A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-07 paper.Rmd
start on revising section 5

## Overview

There are `r 160 * 18`, `r 150 * 18` and `r 133 * 18 - nrow(filter(vi_survey, null_lineup, !attention_check))` lineups evaluation made by 160, 150 and 133 subjects recruited for data collection periods I, II and III respectively. In the total of `r nrow(filter(vi_survey, !attention_check))` lineup evaluations, `r nrow(filter(vi_survey, type == "polynomial", !attention_check))` use lineups produced by the non-linearity model, and `r nrow(filter(vi_survey, type == "heteroskedasticity", !attention_check))` use lineups produced by the heteroskedasticity model. Besides, there are `r nrow(filter(vi_survey, attention_check))` attention checks and `r nrow(filter(vi_survey, null_lineup, !attention_check))` evaluations on null lineups needed for the estimate of $\alpha$ not included in the analysis. The collated dataset is provided in `vi_survey` of the `visage` `R` package.

In the following analysis, lineups with uniform fitted values will be the focus. Visual patterns are more likely to be revealed under a uniform distribution. Additionally, we have collected extra evaluations on these lineups, which will result in a more reliable analysis. Analysis of lineups with other fitted value distributions can be found in Section \ref{effect-of-fitted-value-distributions}.

## Power comparison of the tests

@@

## Comparison of test decisions based on $p$-values

@@

## Effect of amount of non-linearity

@@

## Effect of shape of heteroskedasticity

>>>>

Data collection resulted in a total of `r nrow(filter(vi_survey, !attention_check))` lineup evaluations, `r nrow(filter(vi_survey, type == "polynomial", !attention_check))` for the non-linearity model, and `r nrow(filter(vi_survey, type == "heteroskedasticity", !attention_check))` for the heteroskedasticity model. This was done in three collection periods, called I, II, III, with `r 160 * 18`, `r 150 * 18` and `r 133 * 18 - nrow(filter(vi_survey, null_lineup, !attention_check))` lineups evaluated by 160, 150 and 133 participants, respectively. There were `r nrow(filter(vi_survey, attention_check))` attention check lineups used to filter false participant attempts. To estimate $\alpha$ for calculating statistical significance there were `r nrow(filter(vi_survey, null_lineup, !attention_check))` evaluations of null lineups. Neither the attention checks nor null lineups were used for the subsequent analysis. The de-identified data, `vi_survey`,  is made available in the R package, `visage`.

For clarity and consistency, the analysis in Sections \ref{power-analysis}-\ref{hetero-analysis}, uses only on results from lineups generated with uniform fitted values, a total of `r nrow(filter(vi_survey, x_dist == "uniform"))` lineups and XXX evaluation. More evaluations were made using the uniform fitted value distribution, and it avoids potentially confounding results of the analysis by combining values from different power curves. We also expect that visual patterns are more clearly visible when the fitted value distribution is uniform. Section \ref{effect-of-fitted-value-distributions} contains comparison of the effect of fitted value distribution on power.

## Power comparison of the tests~\label{power-analysis}

@@

## Comparison of test decisions based on $p$-values~\label{p-value}

@@

## Effect of amount of non-linearity~\label{nonlin-analysis}

@@

## Effect of shape of heteroskedasticity~\label{hetero-analysis}

### 2023-04-07 paper.tex
small section heading change
start on revising section 5

\author{\name{Weihao Li$^{a}$, Dianne Cook$^{a}$, Emi
Tanaka$^{a}$, Susan VanderPlas$^{b}$}
\affil{$^{a}$Department of Econometrics and Business Statistics, Monash
University, Clayton, VIC, Australia; $^{b}$Department of Statistics,
University of Nebraska, Lincoln, Nebraska, USA}
}

\thanks{CONTACT Weihao
Li. Email: \href{mailto:weihao.li@monash.edu}{\nolinkurl{weihao.li@monash.edu}}, Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}, Emi
Tanaka. Email: \href{mailto:emi.tanaka@monash.edu}{\nolinkurl{emi.tanaka@monash.edu}}, Susan
VanderPlas. Email: \href{mailto:susan.vanderplas@unl.edu}{\nolinkurl{susan.vanderplas@unl.edu}}}

@@

One may argue that reading diagnostic plots is to some extent subjective
and indecisive compared to those rigorous statistical procedures as it
relies on graphical perception - human ability to interpret and decode
the information embedded in graph \citep{cleveland_graphical_1984}.
Further, the degree of the presence of the visual features typically can
not be measured quantitatively and objectively, which may lead to over
or under-interpretations of the data. For instance, people
over-interpret the separation between gene groups in a two-dimensional
projection from a linear discriminant analysis when in fact there are no
differences in the expression levels between the gene groups and
separation is not an uncommon occurrence
\citep{roy_chowdhury_using_2015}.

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots. Later,
\citet{buja_statistical_2009} proposed the lineup protocol as a visual
test inspired by the ``police lineup'' or ``identity parade'' which is
the act of asking the eyewitness to identify criminal suspect from a
group of irrelevant people. The protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) null plots have the identical graphical procedure except the
data has been replaced with data consistent with \(H_0\), that the model
is correctly specified. Then, an observer who have not seen the data
plot will be asked to point out the most different plot from the lineup.
Under \(H_0\), it is expected that the data plot would have no
distinguishable difference from the null plots, and the probability that
the observer correctly picks the data plot is \(1/m\). If one rejects
\(H_0\) as the observer correctly picks the data plot, then the Type I
error of this test is \(1/m\).

@@

\hypertarget{overview}{%
\subsection{Overview}\label{overview}}

There are 2880, 2700 and 1674 lineups evaluation made by 160, 150 and
133 subjects recruited for data collection periods I, II and III
respectively. In the total of 7974 lineup evaluations, 3744 use lineups
produced by the non-linearity model, and 4230 use lineups produced by
the heteroskedasticity model. Besides, there are 886 attention checks
and 720 evaluations on null lineups needed for the estimate of
\(\alpha\) not included in the analysis. The collated dataset is
provided in \texttt{vi\_survey} of the \texttt{visage} \texttt{R}
package.

In the following analysis, lineups with uniform fitted values will be
the focus. Visual patterns are more likely to be revealed under a
uniform distribution. Additionally, we have collected extra evaluations
on these lineups, which will result in a more reliable analysis.
Analysis of lineups with other fitted value distributions can be found
in Section \ref{effect-of-fitted-value-distributions}.

\hypertarget{power-comparison-of-different-tests}{%
\subsection{Power comparison of different
tests}\label{power-comparison-of-different-tests}}

@@

\hypertarget{comparison-of-test-decisions-based-on-p-values}{%
\subsection{\texorpdfstring{Comparison of test decisions based on
\(p\)-values}{Comparison of test decisions based on p-values}}\label{comparison-of-test-decisions-based-on-p-values}}

@@

\hypertarget{effect-of-amount-of-non-linearity}{%
\subsection{Effect of amount of
non-linearity}\label{effect-of-amount-of-non-linearity}}

@@

\hypertarget{effect-of-shape-of-heteroskedasticity}{%
\subsection{Effect of shape of
heteroskedasticity}\label{effect-of-shape-of-heteroskedasticity}}

>>>>

\author{\name{Weihao Li$^{a}$, Dianne Cook$^{a}$, Emi Tanaka$^{a, b,
c}$, Susan VanderPlas$^{d}$}
\affil{$^{a}$Department of Econometrics and Business Statistics, Monash
University, Clayton, VIC, Australia; $^{b}$Biological Data Science
Institute, Australian National University, Acton, ACT,
Australia; $^{c}$Research School of Finance, Actuarial Studies and
Statistics, Australian National University, Acton, ACT,
Australia; $^{d}$Department of Statistics, University of Nebraska,
Lincoln, Nebraska, USA}
}

\thanks{CONTACT Weihao
Li. Email: \href{mailto:weihao.li@monash.edu}{\nolinkurl{weihao.li@monash.edu}}, Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}, Emi
Tanaka. Email: \href{mailto:emi.tanaka@anu.edu.au}{\nolinkurl{emi.tanaka@anu.edu.au}}, Susan
VanderPlas. Email: \href{mailto:susan.vanderplas@unl.edu}{\nolinkurl{susan.vanderplas@unl.edu}}}

@@

The examination of data plots to infer signal or patterns (or lack
thereof) is fraught with variation in the human ability to interpret and
decode the information embedded in graph
\citep{cleveland_graphical_1984}. Such examination thus feels there is a
subjective nature to evaluate, say, diagnostic plots to infer
appropriateness of statistical models, although arguably we would expect
such evaluation should be done objectively and such human-to-human
variation in ``reading'' data plots should be minimised.

In practice, over-interpretation of a single plot is common. For
instance, in \citet{roy_chowdhury_using_2015}, some over-interpreted the
separation between gene groups in a two-dimensional projection of a
linear discriminant analysis when in fact there are no differences in
the expression levels between the gene groups. To mitigate this
over-interpretation, \citet{buja_statistical_2009} proposed a line-up
protocol to assess plots in a manner analogous to the null hypothesis
significance testing (NHST) framework. More specifically, the protocol
consists of \(m\) randomly placed plots, where one plot is the data
plot, and the remaining \(m - 1\) plots, referred to as the \emph{null
plots}, have the identical graphical procedure as the data plot except
the data is replaced with a data generation mechanism that is consistent
with the null hypothesis, \(H_0\). Then, an observer who have not seen
the data plot will be asked to point out the most different plot from
the lineup. Under \(H_0\), it is expected that the data plot would have
no distinguishable difference from the null plots, and the probability
that the observer correctly picks the data plot is \(1/m\). If one
rejects \(H_0\) as the observer correctly picks the data plot, then the
Type I error of this test is \(1/m\). This protocol requires apriori
specification of \(H_0\) (or at least a null data generating mechanism)
much like the requirement of knowing the distribution of the test
statistic in NHST.

@@

Data collection resulted in a total of 7974 lineup evaluations, 3744 for
the non-linearity model, and 4230 for the heteroskedasticity model. This
was done in three collection periods, called I, II, III, with 2880, 2700
and 1674 lineups evaluated by 160, 150 and 133 participants,
respectively. There were 886 attention check lineups used to filter
false participant attempts. To estimate \(\alpha\) for calculating
statistical significance there were 720 evaluations of null lineups.
Neither the attention checks nor null lineups were used for the
subsequent analysis. The de-identified data, \texttt{vi\_survey}, is
made available in the R package, \texttt{visage}.

For clarity and consistency, the analysis in Sections
\ref{power-analysis}-\ref{hetero-analysis}, uses only on results from
lineups generated with uniform fitted values, a total of 4135 lineups
and XXX evaluation. More evaluations were made using the uniform fitted
value distribution, and it avoids potentially confounding results of the
analysis by combining values from different power curves. We also expect
that visual patterns are more clearly visible when the fitted value
distribution is uniform. Section
\ref{effect-of-fitted-value-distributions} contains comparison of the
effect of fitted value distribution on power.

\hypertarget{power-comparison-of-the-tests}{%
\subsection{\texorpdfstring{Power comparison of the
tests\textasciitilde{}\label{power-analysis}}{Power comparison of the tests\textasciitilde{}}}\label{power-comparison-of-the-tests}}

@@

\hypertarget{comparison-of-test-decisions-based-on-p-values}{%
\subsection{\texorpdfstring{Comparison of test decisions based on
\(p\)-values\textasciitilde{}\label{p-value}}{Comparison of test decisions based on p-values\textasciitilde{}}}\label{comparison-of-test-decisions-based-on-p-values}}

@@

\hypertarget{effect-of-amount-of-non-linearity}{%
\subsection{\texorpdfstring{Effect of amount of
non-linearity\textasciitilde{}\label{nonlin-analysis}}{Effect of amount of non-linearity\textasciitilde{}}}\label{effect-of-amount-of-non-linearity}}

@@

\hypertarget{effect-of-shape-of-heteroskedasticity}{%
\subsection{\texorpdfstring{Effect of shape of
heteroskedasticity\textasciitilde{}\label{hetero-analysis}}{Effect of shape of heteroskedasticity\textasciitilde{}}}\label{effect-of-shape-of-heteroskedasticity}}
